Compositional Quantitative Resource Analysis of C Programs

### 2014-06-24 paper/main.tex
some fixes and started with the description of the logic

We have implemented and proved sound such a quantitative Hoare for
CompCert Clight in the Coq Proof Assistant.  Because we use a shallow
embedding in Coq, every function that can be expressed in Coq can in
principle be used as a resource bound.  We can use the logic to prove
a concrete bound for a given resource metric, or a bound that is
parametric in a set of resource metrics.  It is also possible to
incorporate boolean preconditions into the bounds to express that the
bound is only valid for a certain class of inputs.  To this end, we
allow the potential function $\Phi$ in the pre and post conditions to
take non-negative number or $\infty$ (infinity) as values.  Infinity,
plays the same role as $\bot$ in Hoare logic.  Boolean assertions can
be embedded into potential functions by mapping $\bot$ to $\infty$ and
$\top$ to $0$.

@@

The small-step semantics is standard, except that it tracks the
resource consumption of a program.  The semantics is oblivious to the
resource of interest for the user of our system.  We achieve this
independence by parameterizing evaluations with a resource
\emph{metric} $M$; a tuple of nine rational numbers and one map $M_e :
\text{Expr} \to \Q$ from expressions to rational numbers.
$$
M = (
M_a,
M_b,
M_r,
M_u,
M_f,
M_l,
M_c^1,
M_c^2,
M_q,
M_e(E)
)
$$
Each of these rational numbers indicates the amount of resource
consumed by a corresponding operation.  If one of these numbers is
negative, it means that resources are released by the program.  The
metric that we use in the implementation can also depend on other
information that is statically available such as the name of the called
function or the number of arguments.

@@

\begin{figure}[t!]
\begin{mathpar}
%
\Rule{S:Assert}
{ \code{istrue}~\sem{e}_\state }
{ (\state, \code{assert}~e, \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_a)
}
%
\and
\Rule{S:BrkSeq}
{}
{ (\state, \code{break}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{break}, \cont, \cost)
}
%
\and
\Rule{S:BrkLoop}
{}
{ (\state, \code{break}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_b)
}
%
\and
\Rule{S:RetSeq}
{}
{ (\state, \code{return}~x, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{return}~x, \cont, \cost)
}
\and
%
\and
\Rule{S:RetLoop}
{}
{ (\state, \code{return}~x, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{return}~x, \cont, \cost)
}
%
\and
\Rule{S:RetCall}
{ \state = (\_, \gamma) \\
  \state' = (\theta, \gamma)[r \mapsto \state(x)]
}
{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_r)
}
%
\and
\Rule{S:Update}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
}
%
\and
\Rule{S:Call}
{ \fenv f = (\vec x, S_f) \\
  \state = (\theta, \gamma) \\
  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
}
{ (\state, r \gets f(\vec y), \cont, \cost)
\smallstep{M}
(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
}
%
\and
\Rule{S:Loop}
{}
{ (\state, \code{loop}~S, \cont, \cost)
\smallstep{M}
(\state, S, \Kloop S \cont, \cost)
}
%
\and
\Rule{S:SkipLoop}
{}
{ (\state, \code{skip}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{loop}~S, \cont, \cost {-} M_l)
}
%
\and
\Rule{S:IfTrue}
{ \code{istrue}~\sem{e}_\sigma }
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \cont, \cost {-} M_c^1)
}
%
\and
\Rule{S:IfFalse}
{ \code{isfalse}~\sem{e}_\sigma }
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep{M}
(\state, S_2, \cont, \cost {-} M_c^2)
}
%
\and
\Rule{S:Seq}
{ }
{ (\state, S_1; S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \Kseq {S_2} \cont, \cost)
}
%
\and
\Rule{S:SkipSeq}
{}
{ (\state, \code{skip}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, S, \cont, \cost {-} M_q)
}
%

\section{Proof System}

@@

>>>>

We have implemented and proved sound such a quantitative Hoare for
CompCert Clight in the Coq Proof Assistant.  Because we use a shallow
embedding in Coq, every function that can be expressed in Coq can in
principle be used as a resource bound.  We can use the logic to prove
a concrete bound for a given resource metric, or a bound that is
parametric in a set of resource metrics.  It is also possible to
incorporate boolean preconditions into the bounds to express that the
bound is only valid for a certain class of inputs.  To this end, we
allow the potential function $\Phi$ in the pre and post conditions to
take non-negative number or $\infty$ (infinity) as values.  Infinity,
plays the same role as $\mathit{false}$ in Hoare logic.  Boolean assertions can
be embedded into potential functions by mapping $\mathit{false}$ to $\infty$ and
$\mathit{true}$ to $0$.

@@

The small-step semantics is standard, except that it tracks the
resource consumption of a program.  The semantics is oblivious to the
resource of interest for the user of our system.  We achieve this
independence by parameterizing evaluations with a resource
\emph{metric} $M$; a tuple of nine rational numbers, one map $M_e :
\text{Expr} \to \Q$ from expressions to rational numbers, and one map
$M_t : \Z \to \Q$ from integers to rational numbers.
$$
M = (
M_a,
M_b,
M_r,
M_u,
M_f,
M_l,
M_c^1,
M_c^2,
M_q,
M_e(E),
M_t(n)
)
$$
Each of these rational numbers indicates the amount of resource
consumed by a corresponding operation.  If one of these numbers is
negative, it means that resources are released by the program.  The
metric that we use in the implementation can also depend on other
information that is statically available such as the name of the called
function or the number of arguments.

\begin{figure}[t!]
\begin{mathpar}
%
\Rule{S:Assert}
{ \code{istrue}~\sem{e}_\state }
{ (\state, \code{assert}~e, \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_a)
}
%
\and
\Rule{S:BrkSeq}
{}
{ (\state, \code{break}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{break}, \cont, \cost)
}
%
\and
\Rule{S:BrkLoop}
{}
{ (\state, \code{break}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_b)
}
%
\and
\Rule{S:RetSeq}
{}
{ (\state, \code{return}~x, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{return}~x, \cont, \cost)
}
\and
%
\and
\Rule{S:RetLoop}
{}
{ (\state, \code{return}~x, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{return}~x, \cont, \cost)
}
%
\and
\Rule{S:RetCall}
{ \state = (\_, \gamma) \\
  \state' = (\theta, \gamma)[r \mapsto \state(x)]
}
{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_r)
}
%
\and
\Rule{S:Update}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
}
%
\and
\Rule{S:Call}
{ \fenv f = (\vec x, S_f) \\
  \state = (\theta, \gamma) \\
  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
}
{ (\state, r \gets f(\vec y), \cont, \cost)
\smallstep{M}
(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
}
%
\and
\Rule{S:Loop}
{}
{ (\state, \code{loop}~S, \cont, \cost)
\smallstep{M}
(\state, S, \Kloop S \cont, \cost)
}
%
\and
\Rule{S:SkipLoop}
{}
{ (\state, \code{skip}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{loop}~S, \cont, \cost {-} M_l)
}
%
\and
\Rule{S:IfTrue}
{ \code{istrue}~\sem{e}_\sigma }
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \cont, \cost {-} M_c^1)
}
%
\and
\Rule{S:IfFalse}
{ \code{isfalse}~\sem{e}_\sigma }
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep{M}
(\state, S_2, \cont, \cost {-} M_c^2)
}
%
\and
\Rule{S:Seq}
{ }
{ (\state, S_1; S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \Kseq {S_2} \cont, \cost)
}
%
\and
\Rule{S:SkipSeq}
{}
{ (\state, \code{skip}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, S, \cont, \cost {-} M_q)
}
%
\and
\Rule{S:Tick}
{ }
{ (\state, \code{tick}(n), \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_t(n))
}

@@

\section{Quantitative Hoare Logic}

In this section we describe a simplified version of the quantitative
Hoare logic that we use in Coq to interactively prove resource bounds.

We generalize classic Hoare logic to express not only classical
boolean-valued assertions but also assertions that talk about the
future resource usage.  Instead of the usual assertions $P : \State
\to \mathit{bool}$ of Hoare logic we use assertions
$$
P : \State \to \Q \cup \{ \infty \} \; .
$$
This can be understood as a refinement of boolean assertions where
$\mathit{false}$ is $\infty$ and $\mathit{true}$ is refined by $\Q$.
We write $\Assn$ for $\State \to \Q \cup \{ \infty \}$ and $\bot$ for
$\lambda \state . \, \infty$.  To use Coq's support for propositional
reasoning, assertions have the type $\State \to \Q \to \mathrm{Prop}$
in the implementation.  For a given $\state \in \State$, such an
assertion can be seen as a set $B \subseteq \Q$ of valid bounds.
However, we find the presentation in this article easier to read.

% The continuation-based semantics of a Clight requires that we
% distinguish pre- and postconditions in the logic to account for
% different possible ways to exit a block of code.  This is approach is
% standard in Hoare logics and followed for instance in Appel's
% separation logic for Clight~\cite{AppelLogic}.  Our postconditions
% $$
% Q = (Q^s, Q^b, Q^r) : \Assn \times \Assn \times (\Val \to \Assn)
% $$
% provide one assertion $Q^s$ for the case in which the block is exited by
% fall through, one assertion $Q^b$ if the block is exited by a \code{break}, and a
% function $Q^r$ from values to assertions in case the block is exited by a
% \code{return}.  The function argument in the last case represents the return
% value and the intended meaning is that the resulting assertion is
% guaranteed to hold for every return value.

% Since we have to deal with recursive functions, we also need a
% function context
% $$\Gamma {:} \funs {\to} ((\Val \times \Mem) {\to} \N \cup \{\infty\}) {\times} ((\Val \times \Mem) {\to} \N {\cup} \{\infty\})$$ 
% that maps function names to their specifications, that is, pre- and
% postconditions.  The precondition depends on the value that is passed
% to the function by the caller and the memory.  The postcondition
% depends on the return value and the memory.  For simplicity, we assume
% that a function has only one argument in this article.  In the Coq
% implementation, an arbitrary number of function arguments is allowed.

% In summary, a quantitative Hoare triple has the form
% $$
% \Gamma \vdash \htriple{P}{S}{Q}
% $$
% where $\Gamma$ is a function context, $P : \Assn$ is a precondition,
% $Q : \Assn \times \Assn \times (\Val \to \Assn)$ is a postcondition,
% and $S$ is a statement.

% Intuitively, an assertion can be seen as a \emph{potential function}
% that maps a program state to a non-negative potential.  The potential
% of the precondition $P$ must be sufficient to cover the cost of the
% execution of the statement $S$ and the potential $Q$ after the execution
% of $S$ (as in amortized resource analysis~\cite{HoffmannAH12}).

@@

\section{Automatic Amortized Analysis}
